Simulations of Ballot Polling Risk-Limiting Audits

### 2022-10-12 providence/sections/Motivation_old.tex

>>>>

\begin{definition}
Let $d \in \{0, a\}$ stand for the null or the alternative hypothesis respectively.
Let $s_{i, d, [n_1, \ldots, n_r]}$ be the probability of observing $i$ ballots for the winner after drawing the $r^{th}$ round of an audit with round schedule $[n_1, \ldots, n_r]$ conditioned on hypothesis $H_d$ \textit{i.e.,} 
\[
s_{i, d, [n_1, \ldots, n_r]} = P\left[K_r = i | (n_1, \ldots, n_r), H_d\right]. 
%  = s_{i, d, [n_r]},
\]
For $r = 1$ 
\[
 s_{i, d, [n_r]} = \mathsf{Bin}(n_r, p_d, i).
\]
\end{definition}

\begin{example}[Adaptive round size selection] 
Let us assume a malicious auditor with the following strategy for \Minerva audit:
\begin{itemize}
 \item select second round size $n_2$ if $k_{1} = k^* = k^{p_a, p_0, \alpha}_{min, 1} - 1$,
 \item select round size $n'_2$ for $k_1 \leq k^{p_a, p_0, \alpha}_{min, 1} - 2$.
\end{itemize}
This leads to the situation where the true value of $s_{i, d, [n_1, n_2]}$ is missing any contributions from $K_1=k^*$.

\[
 s^{k^*, up}_{i, d, [n_1, n_{2}]} = \sum_{j = x}^{\min\{i, k^{p_a, p_0, \alpha}_{min, 1}-1\}} Bin(n_{2} - n_1, p_d, i - j) \cdot s_{j, d, [n_1]},
\]

\[
 s^{k^*, down}_{i, d, [n_1, n'_{2}]} = \sum_{j = \max\{0, i - (n'_2 - n_1)\}}^{k^*-1} Bin(n'_{2} - n_1, p_d, i - j) \cdot s_{j, d, [n_1]},
\]

One would need to to prove that for $n_2, n'_2$ and for any choice of $k^*$ the following holds:
\[
\sum_{i = k^{p_a, p_0, \alpha}_{min, 2, [n_1, n_2]}}^{n_r} s^{k^*, up}_{i, 0, [n_1, n_2]} \le \alpha \sum_{i = k^{p_a, p_0, \alpha}_{min, 2, [n_1, n_2]}}^{n_r} s^{k^*}_{i, A, [n_1, n_2]}.
\]
\end{example}

### 2022-10-12 providence/sections/adversary.tex

To illustrate the practical implication of this property, we consider a toy example: an RLA of a two-candidate contest with margin $0.01$ and risk limit $0.1$. 
Suppose we wish to achieve a conditional stopping probability $0.9$ in each round of the audit. For \Providence, we can compute a new round size for each round based on the previous samples. For \Minerva, however, we would have a predetermined round schedule. We use the default \Minerva round schedule of audit software Arlo \cite{arlo} (used by most states performing an RLA), which is is $[x, 2.5x, 6.25x, ...]$; that is, the next marginal round size is $1.5$ times the current one. This multiplier of $1.5$ is known to give, over a wide range of margins, a probability of stopping roughly $0.9$ in the second round if the first round size has probability of stopping $0.9$.

>>>>

To illustrate the practical implication of this property, we consider a toy example: an RLA of a two-candidate contest with margin $1\%$ and risk limit $10\%$. 
Suppose we wish to achieve a conditional stopping probability $0.9$ in each round of the audit. For \Providence, we can compute a new round size for each round based on the previous samples. For \Minerva, however, we would have a predetermined round schedule. We use the default \Minerva round schedule of audit software Arlo \cite{arlo} (used by most states performing an RLA), which is is $[x, 2.5x, 6.25x, ...]$; that is, the next marginal round size is $1.5$ times the current one. This multiplier of $1.5$ is known to give, over a wide range of margins, a probability of stopping roughly $0.9$ in the second round if the first round size has probability of stopping $0.9$.

### 2022-10-12 providence/sections/appendix-copy.tex

\begin{lemma}
    \label{lemma:sigma_increasing}
For $0<p_0< p_a< 1$ and $n>0$, the ratio $\sigma(k,p_a,p_0,n)$ is strictly increasing as a function of $k$ for $0\le k\le n$.
\end{lemma}
\begin{proof}
See \cite[Lemma 4]{usenix_minerva}. 
\end{proof}

@@

\begin{proof}
See \cite[Lemma 2]{usenix_minerva}. 
\end{proof}

\begin{lemma}
    \label{lemma:tau1_increasing}
For $0<p_0< p_a< 1$ and $n>0$, the ratio $\tau_1(k,p_a,p_0,n)$ is strictly increasing as a function of $k$ for $0\le k\le n$.
\end{lemma}
\begin{proof}
    Apply Lemmas \ref{lemma:sigma_increasing}-\ref{lemma:frac_sums_increasing}.
\end{proof}
\begin{lemma}
    \label{lemma:imin-exists}
    Given a strictly monotone increasing sequence: $x_1, x_2, \ldots x_n $ and some constant $A$,
    %there exists an $i_{min}$ such that
    $$A \le x_i \Leftrightarrow \exists i_{min} \le i ~\text{s.t.}~   x_{i_{min} -1} < A \le x_{i_{min}} \le x_{i},$$
    unless $A\le x_1$, in which case $i_{min} =1 $.
\end{lemma}
\begin{proof}
    Evident.
\end{proof}

\begin{lemma}
    \label{lemma:minerva2_kmin_exists}
    For $\mathcal{A}=(\alpha,p_a, p_0,k_{j-1},n_{j-1},n_j)$-\Providence, there exists\\ a 
    $k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  = 
    k_{min,j}(\Providence, p_a, p_0, k_{j-1}, n_{j-1}, n_j)$ such that $$\mathcal{A}(X_j)=\text{Correct}\iff k_j\ge k_{min,j}(\Providence,  \bm{n_j}, p_a, p_0).$$
\end{lemma}
\begin{proof}
    From Definition~\ref{def:minervatwo}, $$\mathcal{A}(X_j)=\text{Correct}\iff \omega_j(k_{j}, k_{j-1}, p_a, p_0, n_j, n_{j-1}) \ge \frac{1}{\alpha}.$$
    Now to apply Lemma~\ref{lemma:imin-exists}, it suffices to show that
    $\omega_j$ is monotone increasing with respect to $k_j$.
    For $j=1$, we have $\omega_1=\tau_1$, so $\omega_1$ is strictly increasing by Lemma \ref{lemma:tau1_increasing}. For $j\ge 2$,
    $$\omega_j(k_j,k_{j-1},p_a,p_0,n_j,n_{j-1},\alpha)=$$$$\sigma(k_{j-1},p_a,p_0,n_{j-1})\cdot \tau_1(k_{j}-k_{j-1},p_a,p_0,n_j-n_{j-1}).$$
    As a function of $k_j$, $\sigma$ is constant, and thus $\omega$ is strictly increasing by Lemma \ref{lemma:tau1_increasing}. Therefore by Lemma \ref{lemma:imin-exists}, we have the desired property.
\end{proof}

@@

\begin{definition} 
\label{def:kmin}
Let $[n_1, \ldots, n_j]$ be the round schedule of an audit that has not stopped by the round $j-1$. Let us define 
\begin{small}
\begin{equation}\label{eq:kMin}
k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  =
  \min\left\{k : \omega_j(k, k_{j-1},p_a,p_0,n_j, n_{j-1}) \geq \frac{1}{\alpha}  \right\}.
%  \min\left\{k : \sigma(k_{r-1},p_a,p_0,n_{r-1})\cdot \tau_1(k-k_{r-1},p_a,p_0,n_r-n_{r-1}) \geq \frac{1}{\alpha}  \right\}$
\end{equation}
\end{small}
\end{definition}
As we have seen in Lemma \label{lemma:minerva2_kmin_exists}, such a value of $k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}$ exists and $k_j \geq k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} $ if and only if the result of the audit is Correct, (\textit{i.e.,} the stopping condition in Definition~\ref{def:minervatwo} holds).

The following lemma shows a Markov-like property of \Providence audit (\textit{i.e.,}
for an audit that has not stopped in the first $j-1$ rounds, only cumulative results of the round $j-1$ matter: cumulative sample size $n_{j-1}$ and the number of ballots for the winner $k_{j-1}$).

\begin{lemma}\label{lemma:markov}
Let $[n_1, \ldots, n_{j-1}, n_j]$ be a round schedule for an execution of  \Providence audit that has not stopped
in any of its first $j-1$ rounds (\textit{i.e.,} for every $i = 1, \ldots, j-1$:
$k_i < k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} $), then:

\[ 
k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} = k^{p_a, p_0, \alpha, k_{j-1}}_{min, 2, n_{j-1}, n_j} .
\]
\end{lemma}
% \fpo{this can be used to prove that \Providence is more efficient than \Minerva and \BRAVO}
\begin{proof}
Let $k_{j-1}$ denote the number of ballots drawn for the declared winner up to the round $j-1$ (out of $n_{j-1}$ sampled ballots). The stopping decision for the round $j$ is made as follows:

\[
 k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  = \min\left\{k : \omega_{j}(k, k_{r-1}, p_a, p_0, n_r, n_{r-1}) \geq \frac{1}{\alpha}  \right\} = 
\]
\[
  =  k^{p_a, p_0, \alpha, k_{j-1}}_{min, 2, n_{j-1}, n_j}  
\]

@@

That is, the stopping condition is equivalent to that of a two round audit with the same cumulative votes for the winner and cumulative round sizes: the first round is of size $n_{j-1}$ and has $k_{j-1}$ votes for the winner, and the second (cumulative) round size is $n_j$ with $k_j$ (cumulative) votes for the winner. Compare this to the similar property for the $\Bravo$ stopping condition.

>>>>

\begin{lemma}
    \label{lemma:sigma_increasing}
For $0<p_0< p_a< 1$ and $n>0$, the ratio $\sigma(k,p_a,p_0,n)$ is strictly increasing as a function of $k$ for $0\le k\le n$.
\end{lemma}
\begin{proof}
From $0<p_0< p_a< 1$, we get
$$
\frac{p_a}{p_0}>1
$$
and
$$
1-p_0>1-p_a\implies \frac{1-p_0}{1-p_a}>1,
$$
and thus
$$
\frac{p_a(1-p_0)}{p_0(1-p_a)}>1.
$$
Now simply observe that
$$
\frac{p_a(1-p_0)}{p_0(1-p_a)}\cdot\sigma(k, p_a, p_0, n)
=\frac{p_a(1-p_0)}{p_0(1-p_a)}\cdot\frac{p_a^{k} (1-p_a)^{n-k}}{p_a^{k} (1-p_a)^{n-k}}
$$$$
=\frac{p_a^{k+1} (1-p_a)^{n-(k+1)}}{p_a^{k+1} (1-p_a)^{n-(k+1)}}
=\sigma(k+1, p_a, p_0, n).
$$
\end{proof}

@@

\begin{proof}
    Note that $z_i$ is a weighted average of the values of $\frac{a_j}{b_j}$ for $j \geq i$:
    $$z_i = \sum_{j=i}^n y_j \frac{a_j}{b_j}$$  ~for~ $$y_j = \frac{b_j}{\sum _{j=i}^n b_j} > 0.$$
    Further, $\sum _{j=i}^n y_j = 1$ and hence  $y_j \leq 1$ ~and~ $y_j = 1 \Leftrightarrow i=j=n$.
    Observe that, because $\frac{a_i}{b_i}$ is monotone increasing,
    $z_i \geq \frac{a_i}{b_i}$
    with equality if and only if $i=n$.
    Suppose $i < n$. Then
    $$z_{i+1} \geq \frac{a_{i+1}}{b_{i+1}} > \frac{a_i}{b_i}, $$
    and
    $$z_i = y_i \frac{a_i}{b_i} + (1-y_i) z_{i+1} < z_{i+1}.$$
    Thus $z_i$ is also monotone increasing.
\end{proof}

\begin{lemma}
    \label{lemma:tau1_increasing}
For $0<p_0< p_a< 1$ and $n>0$, the ratio $\tau_1(k,p_a,p_0,n)$ is strictly increasing as a function $k$ for $0\le k\le n$.
\end{lemma}
\begin{proof}
    Apply Lemmas \ref{lemma:sigma_increasing}-\ref{lemma:frac_sums_increasing}.
\end{proof}
\begin{lemma}
    \label{lemma:imin-exists}
    Given a strictly monotone increasing sequence: $x_1, x_2, \ldots x_n $ and some constant $A$,
    %there exists an $i_{min}$ such that
    $$A \le x_i \Leftrightarrow \exists i_{min} \le i ~\text{s.t.}~   x_{i_{min} -1} < A \le x_{i_{min}} \le x_{i},$$
    unless $A\le x_1$, in which case $i_{min} =1 $.
\end{lemma}
\begin{proof}
    Evident.
\end{proof}

\begin{lemma}
    \label{lemma:minerva2_kmin_exists}
    For $\mathcal{A}=(\alpha,p_a, p_0,k_{j-1},n_{j-1},n_j)$-\Providence, there exists\\ a 
    $k_{min, j, n_{j-1}, n_j}^{p_a, p_0, k_{j-1}} = 
    k_{min,j}(\Providence, p_a, p_0, k_{j-1}, n_{j-1}, n_j)$ such that $$\mathcal{A}(X_j)=\text{Correct}\iff k_j\ge k_{min,j}(\Providence,  \bm{n_j}, p_a, p_0).$$
\end{lemma}
\begin{proof}
    From Definition~\ref{def:minervatwo}, $$\mathcal{A}(X_j)=\text{Correct}\iff \omega_j(k_{j}, k_{j-1}, p_a, p_0, n_j, n_{j-1}) \ge \frac{1}{\alpha}.$$
    Now to apply Lemma~\ref{lemma:imin-exists}, it suffices to show that
    $\omega_j$ is monotone increasing with respect to $k_j$.
    For $j=1$, we have $\omega_1=\tau_1$, so $\omega_1$ is strictly increasing by Lemma \ref{lemma:tau1_increasing}. For $j\ge 2$,
    $$\omega_j(k_j,k_{j-1},p_a,p_0,n_j,n_{j-1},\alpha)=$$$$\sigma(k_{j-1},p_a,p_0,n_{j-1})\cdot \tau_1(k_{j}-k_{j-1},p_a,p_0,n_j-n_{j-1}).$$
    As a function of $k_j$, $\sigma$ is constant, and thus $\omega$ is strictly increasing by Lemma \ref{lemma:tau1_increasing}. Therefore by Lemma \ref{lemma:imin-exists}, we have the desired property.
\end{proof}

@@

\begin{definition} 
Let $[n_1, \ldots, n_r]$ be the round schedule of an audit that has not stopped by the round $r-1$. Let us define 
\begin{small}
\begin{equation}\label{eq:kMin}
k^{p_a, p_0, \alpha, k_{r-1}}_{min, r, [n_1, \ldots, n_r]} =
  \min\left\{k : \omega_r(k, k_{r-1},p_a,p_0,n_r, n_{r-1}) \geq \frac{1}{\alpha}  \right\}.
%  \min\left\{k : \sigma(k_{r-1},p_a,p_0,n_{r-1})\cdot \tau_1(k-k_{r-1},p_a,p_0,n_r-n_{r-1}) \geq \frac{1}{\alpha}  \right\}$
\end{equation}
\end{small}
Then if $k_r \geq k^{p_a, p_0, \alpha, k_{r-1}}_{min, r, [n_1, \ldots, n_r]}$ then the result of the audit is Correct (\textit{i.e.,} stopping condition in Definition~\ref{def:minervatwo} holds).
\end{definition}

The following lemma shows a Markov-like property of \Providence audit (\textit{i.e.,}
for an audit that has not stopped in the first $r-1$ rounds, results of the round $r-1$ matter: cumulative sample size $n_{j-1}$ and the number of ballots for the winner $k_{r-1}$).

\begin{lemma}\label{lemma:markov}
Let $[n_1, \ldots, n_{r-1}, n_r]$ be a round schedule for an execution of  \Providence audit that has not stopped
in any of its first $r-1$ rounds (\textit{i.e.,} for every $j = 1, \ldots, r-1$:
$k_j < k^{p_a, p_0, \alpha}_{min, r, [n_1, \ldots, n_j]}$), then:

\[ 
k^{p_a, p_0, \alpha, k_{r-1}}_{min, [n_1, \ldots, n_{r-1}, n_r]} = k^{p_a, p_0, \alpha, k_{r-1}}_{min, [n_{r-1}, n_r]}.
\]
\end{lemma}
% \fpo{this can be used to prove that \Providence is more efficient than \Minerva and \BRAVO}
\begin{proof}
Let $k_{r-1}$ denote the number of ballots drawn for the declared winner up to the round $r-1$ (out of $n_{r-1}$ sampled ballots). The stopping decision for the round $r$ is made as follows:

\[
 k^{p_a, p_0, \alpha, k_{r-1}}_{min, r, [n_1, \ldots, n_r]} = \min\left\{k : \omega_{j}(k, k_{r-1}, p_a, p_0, n_r, n_{r-1}) \geq \frac{1}{\alpha}  \right\} = 
\]
\[
  =  k^{p_a, p_0, \alpha, k_{r-1}}_{min, 2, [n_{r-1}, n_r]}
\]

@@

### 2022-10-12 providence/sections/intro.tex

\subsection{Open Questions in the Literature}
A major limitation of \Minerva is that one needs to determine the round schedule before the audit begins because \Minerva has not been shown to be risk-limiting if an adversary can choose subsequent round sizes after knowing the sample drawn. \BRAVO, on the other hand, is not limited in this manner. This allows \BRAVO audits the flexibility of choosing smaller subsequent round sizes if the sample drawn so far is a ``good'' sample. An open question is whether a ballot polling RLA exists with the efficiency of \Minerva and this flexibility of \BRAVO.

>>>>

\subsection{Open Questions in the Literature}
A major limitation of \Minerva is that one needs to determine the round schedule before the audit begins, because \Minerva has not been shown to be risk-limiting if an adversary can choose subsequent round sizes after knowing the sample drawn. \BRAVO, on the other hand, is not limited in this manner. This allows \BRAVO audits the flexibility of choosing smaller subsequent round sizes if the sample drawn so far is a ``good'' sample. An open question is whether a ballot polling RLA exists with the efficiency of \Minerva and this flexibility of \BRAVO.

### 2022-10-12 providence/sections/sims.tex

% simulations
We use simulations to provide additional evidence for theoretical claims and to gain insight into audit behavior. As done in \cite{simulations}, we use margins from the 2020 US Presidential election---state-wide pairwise margins of 5\% or more between the two leading candidates. Narrower margins are computationally expensive, especially for the simulations of tied elections, which, by design, have a low probability of stopping and hence quickly increase in sample size. We use the simulator in the R2B2 software library\cite{r2b2_anon}. For each margin, for each hypothesis, we perform $10^4$ trials. All trials use a $10\%$ risk limit, a maximum of $5$ rounds, and a conditional stopping probability of $0.90$ in each round. That is, each subsequent round size is selected to be large enough, assuming the announced tally is correct and given the tally of previous rounds, to give a $0.90$ probability of stopping in the current round.

@@

We now investigate the efficiency of \Providence compared to \Minerva, SO \BRAVO, and EoR \BRAVO by taking a single contest as an example, Texas with margin $5.7\%$. We run an additional $10^4$ simulations for each of the three other audits assuming the announced margin for comparison. Both \BRAVO implementations use a conditional stopping probability of $0.9$ for each round, while \Minerva uses a first round size with stopping probability $0.9$ and a multiplier of $1.5$ to obtain subsequent round sizes. Figure~\ref{fig:prov-asn} shows the probability of stopping as a function of the number of ballots sampled, a plot similar to those presented in \cite{simulations}. Points above (higher probability of stopping) and to the left (fewer ballots) represent more efficient audits. As shown, \Providence has comparable efficiency to \Minerva, while both are significantly more efficient than either implementation of \BRAVO. In a contest with a narrow margin (in the 2020 US Presidential election, eight states had margins less than $3\%$) the difference in number of ballots sampled could correspond to many days of work before a certification deadline.
% Section~\ref{sec:workload} discusses workload in more depth.

>>>>

% simulations
We use simulations to provide additional evidence for theoretical claims and to gain insight into audit behavior. As in \cite{simulations}, we use margins from the 2020 US Presidential election---state-wide pairwise margins of 5\% or more between the two leading candidates. Narrower margins are computationally expensive, especially for the simulations of tied elections, which, by design, have a low probability of stopping and hence quickly increase in sample size. We use the simulator in the R2B2 software library\cite{r2b2_anon}. For each margin, for each hypothesis, we perform $10^4$ trials. All trials use a $10\%$ risk limit, a maximum of $5$ rounds, and a conditional stopping probability of $0.90$ in each round. That is, each subsequent round size is selected to be large enough, assuming the announced tally is correct and given the tally of previous rounds, to give a $0.90$ probability of stopping in the current round.

@@

Figure~\ref{fig:prov-asn} plots the probability of stopping as a function of the number of ballots sampled. Points above (higher probability of stopping) and to the left (fewer ballots) represent more efficient audits. As shown, \Providence has comparable efficiency to \Minerva, while both are significantly more efficient than either implementation of \BRAVO. In a contest with a narrow margin (in the 2020 US Presidential election, eight states had margins less than $3\%$) the difference in number of ballots sampled could correspond to many days of work. 
% Section~\ref{sec:workload} discusses workload in more depth.
